{\scshape Denicek}: Computational Substrate for\\ Document-Oriented End-User Programming

### 2024-12-13 paper.tex
Tweaks, updates in walkthrough and substrate

\begin{itemize}
\item \emph{Programming by Demonstration.} Allow users to construct simple programs by
  demonstrating the steps of the expected behavior using concrete examples~\cite{leiva-2021-rapido,cypher-1993-pbd}.
\item \emph{Local-First Collaboration.} Multiple users should be able to concurrently
  modify a single document and merge their changes, preferrably without requiring a central server~\cite{kleppmann-2019-local,litt-2022-peritext}.
\item \emph{End-User Debugging.} The user should be able to ask provenance questions~\cite{cheney-2009-provenance} --
  why a computation resulted in a particular value and what inputs contributed to the result \cite{ko-2009-whyline}.
\item \emph{Schema Change Control.} When the user evolves the structure of the document, affected
  data and formulas should automatically co-evolve to match the new structure~\cite{litt-2020-cambria,edwards-2025-schema}.
\item \emph{Concrete Programming.} It should be possible to reuse parts of program logic, or formulas,
  without introducing abstractions, that is, program against concrete values \cite{edwards-2006-copypaste,edwards-2022-copypaste}.
\end{itemize}

\subsubsection*{Two-Stage Methodology}
Our approach is captured by the interior mode \cite{adam-2021-dsr} of design science research.
To \emph{design} Denicek, we identify a number of \emph{formative examples} -- simple programming
tasks that manifest one or more of the desired programming experiences.
Using those examples, we co-design the Denicek substrate and Webnicek (Section~\ref{sec:walk}),
a simple web-based end-user programming environment based on the substrate. Although Webnicek can
be used to complete end-user programming tasks, it is optimized for testing the underlying
substrate rather than for usability.

To \emph{evaluate} Denicek, we use the substrate as the basis of Datnicek, an end-user data
exploration environment inspired by existing systems~\cite{kandel-2011-wrangler,drossos-2020-wrex}.
We use the second stage to evaluate suitability of the Denicek substrate for the development of
end-user programming systems and report the results of our evaluation (Section~\ref{sec:eval}).

@@

@@

\newpage
~
% ==================================================================================================

@@

To providea an overview of the programming experiences supported by the Denicek
substrate, we use a web-based implementation of Denicek with a structure
editor that supports navigation in document and issuing of edit commands. The editor
implements support for a range of programming experiences, discussed in \S\ref{sec:impl}.

\subsubsection*{\circled{A} Adding a Speaker}
The user starts with an empty document, which is represented as a record. They add a field for
each heading and a list {\small\Verb_<ul>_} as a field named {\small\Verb_speakers_}. They use the command
toolbox to add the first speaker.

\subsubsection*{\circled{B} Creating a User Interface} To simplify adding of
further speakers, the user creates a textbox and a button. They enter new speaker
details, construct a temporary {\small\Verb_<li>_} element outside of the main list, copy the value
from the textbox into the new element using source view, and then copy the new element into the
{\small\Verb_speakers_} list.

@@

\subsubsection*{\circled{E} Merging Edits} The two document versions can be automatically merged.
The refactoring is applied to all existing speakers. New speakers added using the ``Add speaker!''
button are also automatically turned into the new format.

\keyideabox{\faLightbulbO}{Collaborative Editing and Interactivity}{Denicek's merging
reapplies edits from another branch on top of the current history (\S\ref{sec:impl-collab}).
Merging is asymmetic, but the order does not matter here. The same merging operation is used
when handling user interaction (\S\ref{sec:impl-interaction}).}

@@

\keyideabox{\faLightbulbO}{Schema Code Co-Evolution}{References are understood
by the substrate and so they are updated by structural edits (\S\ref{sec:impl-schema}). The
evaluation mechanism can replace formulas with values, but also augment them to keep an evaluation
trace for provenance analysis (\S\ref{sec:impl-provenance}).}

@@

\section{The Denicek Substrate}
\label{sec:system}
Denicek represents programs as sequences of edits that construct and transform a computational
document. In this section, we describe the structure of documents and edits, as well as the
operations that form the backbone of the system and are used to implement a range of
end-user programming experiences (\S\ref{sec:impl}).

\subsection{Selectors, Documents and Edits}
A computational document is a tree, consisting of four kinds of nodes (Fig.~\ref{fig:doc}).
References to document locations can be relative or absolute. Both kinds can appear in a
document (reference nodes) and absolute references are used in edits (to denote a target node).
A reference is represented as a sequence of selectors (Fig.~\ref{fig:doc}).
The document model assumes that lists are homogeneous and records heterogeneous, and so the
\ident{Any} selector makes it possible to refer to all children of a list, but there is no
way to refer to all children of a record.
%
Note that Denicek does not use implicit numerical indices for lists and indices of new items have
to be supplied explicitly. The design choice simplifies merging and is discussed later
(\S\ref{sec:discuss}).

\subsubsection*{Document Edits}
The supported document edits and their behavior are listed in Fig.~\ref{fig:edits}. All edits
require \emph{target} to which they are applied. Target is a reference and can contain the
\ident{Any} selector, in which case the edit is applied to multiple nodes simultaneously.
Most edits can only be applied to target node(s) of a particular kind.

The edits are designed to allow any transformation of a document
through a series of steps whose effect can be tracked by the substrate. As illustrated earlier,
Denicek updates references when document structure changes. Fig.~\ref{fig:edits} distinguishes
between edits that keep existing references in a document unchanged (above) and edits that affect
references (below). When a selector is \emph{invalidated}, e.g., when deleting a field or a
list item to which there is a reference, Denicek rejects the edit. Copying also invalidates
selectors because it is ambiguous whether selectors referring to the original location should
refer to the source or the target of the copying after the edit (and references cannot refer to
multiple structurally incompatible locations). Updating a field, wrapping, reordering or deleting
a list item requires updating references in the document correspondingly.

@@

\begin{figure}
\newcommand{\ektablecol}[6]{
\raisebox{-0.2em}{#1} & \sffamily\small{\bfseries #2}\,\;\textit{\footnotesize target, #3} & \sffamily\footnotesize #4 & \sffamily\footnotesize #5 \\[-0.2em]
&\multicolumn{3}{l|}{\sffamily\footnotesize #6}\\[0.3em]
}
\begin{tabular}{|cp{13em}rr|}
\hline
\rowcolor{ekgray}
&&&\\[-1em]
\rowcolor{ekgray}
 & \sffamily\small{\bfseries Edit}\;\,\textit{\footnotesize arguments} & \sffamily\footnotesize\bfseries Target & \sffamily\footnotesize\bfseries Selectors \\[0.2em]
\hline
&&&\\[-1em]
\ektablecol{\faPlus}{Add}{field, node}{Record}{Unchanged}
  {Add \textit{node} as a \textit{field} to the specified record.}
\ektablecol{\faAt}{Append}{node}{List}{Unchanged}
  {Append \textit{node} to the end of the specified list.}
\ektablecol{\faPaperclip}{AppendFrom}{selectors}{List}{Unchanged}
  {Append node from \emph{selectors} to the end of the specified list.}
\ektablecol{\faICursor}{PrimitiveEdit}{transform}{Primitive}{Unchanged}
  {Apply primitive \textit{transform} to the specified primitive.}
\ektablecol{\faCode}{UpdateTag}{old tag, new tag}{Tagged}{Unchanged}
  {Change the tag of a specified list or record from \textit{old} to \textit{new}.}
\hline
&&&\\[-1em]
\ektablecol{\faFont}{UpdateField}{old field, new field}{Record}{Update}
  {Rename the field of a specified record from \textit{old} to \textit{new}.}
\ektablecol{\faTimesCircle}{DeleteField}{field}{Record}{Invalidate}
  {Delete the field \textit{field} of a specified record.}
\ektablecol{\faMinusCircle}{DeleteItem}{index}{List}{Update}
  {Delete the item at a given \textit{index} of a specified list.}
\ektablecol{\faSort}{Reorder}{permutation}{List}{Update}
  {Reorder items of a specified list according to a \textit{permutation}.}
\ektablecol{\faFileO}{WrapRecord}{tag, field}{Any}{Update}
  {Wrap the specified node as a \textit{field} of a new record with \textit{tag}.}
\ektablecol{\faListUl}{WrapList}{tag}{Any}{Update}
  {Wrap the specified node as a sole element of a new list with \textit{tag}.}
\ektablecol{\faCopy}{Copy}{selectors}{Any}{Invalidate}
  {Copy nodes(s) from \textit{selectors}, replacing the specified target(s). }
\hline
\end{tabular}
\vspace{-0.5em}
\caption{Summary of document edit types in Denicek}
\label{fig:edits}
\vspace{-1em}
\end{figure}

@@

\subsubsection*{Automatic Reference Update}
There are two situations in which automatic update of references is undesirable. If an edit is
applied to a singular element of a list (reference contains the \ident{Index} selector),
references that refer into any element of the list should be unchanged.  Such edits may turn
document into an inconsistent state, but they typically do so temporarily during document
construction (checking such edits is discussed in \S\ref{sec:discuss-future}).

Documents can also contain values that can be of multiple different kinds (i.e., a union type).
In such case, references should not be updated when the kind of the value changes. For example,
a formula may be either unevaluated or evaluated. As discussed in \S\ref{sec:impl-eval}, evaluation
involves wrapping, but this should not affect references to the formula.
To support those cases, it is possible to annotate edits that normally affect selectors
(Fig.~\ref{fig:edits}, below) as non-structural. This annotation is required when the
target reference contains the \emph{Index} selector.

\subsubsection*{No Conditional Edits.}
An important aspect of the design is that the effect an edit has on references inside
the document does not depend on the current value of the document. This makes it possible to
define merging solely in terms of edits, without reference to current document state.
(The effect of \ident{Copy} depends only on the existing document structure, but not on its value;
while \ident{AppendFrom} affects only single list element and is thus labelled as a
non-structural edit.)

This design choice makes it impossible to encode computational logic directly in the edits
(e.g., through conditional edits). As we will see in \S\ref{sec:impl-interaction}, such logic
has to be provided as an additional  mechanism on top of the underlying Denicek substrate.

\subsection{Primitive Operations}
\label{sec:system-ops}
Denicek provides three primitive operations. A sequence of edits can be applied to a document, two
sequences of edits can be checked for conflicts or merged. Denicek identifies edit histories by
a (git-like) hash, computed from the hash of the parent and the current edit, which is used to identify
common shared part of the history during conflict checking and merging.

\subsubsection*{Applying Edits}
When applying an edit, Denicek locates the target node and transform it according to the
edit. If the edit may affect references in the document (Fig~\ref{fig:edits}, below), Denicek
updates matching references in the document according to the rules shown in Fig.~\ref{fig:updates}
provided that (i) the edit is not explicitly marked as non-structural, and (ii) the target
reference does not contain \ident{Index} selector, i.e., it targets all elements of any involved
lists. Also note that matching references in the document may be more specific than the edit
target. For example, if we rename {\small\Verb|old|} to {\small\Verb|new|} at {\small\Verb|/foo/*|},
a reference {\small\Verb|/foo/3/old|} will become {\small\Verb|/foo/3/new|}. If the document
contains a reference that would be invalidated by an edit, the edit is rejected.

@@

\begin{figure}
\newcommand{\tttablecol}[5]{
\small{\bfseries #1}\;\,\footnotesize\textit{#2}\,\; --\,\; #5\\[-0.1em]
\quad \footnotesize #3 \;\;$\Rightarrow$\;\; #4 \\[0.3em]
}
\begin{tabular}{|l|}
\hline
\\[-1em]
\tttablecol{UpdateField}{target, old field, new field}{\Verb|/target/old\_field/nested|}{\Verb|/target/new\_field/nested|}
  {Replace Field for matching references.}
\tttablecol{DeleteItem}{target, index}{\Verb|/target/n/nested|}{\Verb|/target/(n-1)/nested|}
  {Decrement Index greater than $n$ in matching refs.}
\tttablecol{Reorder}{target, permutation}{\Verb|/target/n/nested|}{\Verb|/target/permutation(n)/nested|}
  {Update Index using permutation in matching refs.}
\tttablecol{WrapRecord}{target, tag, field}{\Verb|/target/nested|}{\Verb|/target/field/nested|}
  {Insert extra Field selector after matching prefix.}
\tttablecol{WrapList}{target, tag}{\Verb|/target/nested|}{\Verb|/target/*/nested|}
  {Insert extra All selector after matching prefix.}
\hline
\end{tabular}
\vspace{-0.5em}
\caption{How document edits transform references}
\label{fig:updates}
\vspace{-1em}
\end{figure}

@@

maybe

>>>>

\begin{itemize}
\item \emph{Local-First Collaboration.} Multiple users should be able to concurrently
  modify a single document and merge their changes, preferrably without requiring a central server~\cite{kleppmann-2019-local,litt-2022-peritext}.
\item \emph{Programming by Demonstration.} Allow users to construct simple programs by
  demonstrating the steps of the expected behavior using concrete examples~\cite{leiva-2021-rapido,cypher-1993-pbd}.
\item \emph{Incremental Recomputation.} When a part of a document changes, formulas whose result
  depends on it are invalidated and, possibly, automatically recomputed \cite{mcdirmid-2013-usable,horowitz-2023-engraft,petricek-2020-live}.
\item \emph{Schema Change Control.} When the user evolves the structure of the document, affected
  data and formulas should automatically co-evolve to match the new structure~\cite{litt-2020-cambria,edwards-2025-schema}.
\item \emph{End-User Debugging.} The user should be able to ask provenance questions~\cite{cheney-2009-provenance} --
  why a computation resulted in a particular value and what inputs contributed to the result \cite{ko-2009-whyline}.
\item \emph{Concrete Programming.} It should be possible to reuse parts of program logic, or formulas,
  without introducing abstractions, that is, program against concrete values \cite{edwards-2006-copypaste,edwards-2022-copypaste}.
\end{itemize}

\subsubsection*{Two-Stage Methodology}
Our approach is captured by the interior mode \cite{adam-2021-dsr} of design science research.
To \emph{design} Denicek, we identify a number of \emph{formative examples} -- simple programming
tasks that manifest one or more of the desired programming experiences (\S\ref{app:examples}).
Using those examples, we co-design the Denicek substrate and Webnicek (\S\ref{sec:walk}),
a simple web-based end-user programming environment based on the substrate. Although Webnicek can
be used to complete end-user programming tasks, it is optimized for testing the underlying
substrate rather than for usability.

To \emph{evaluate} Denicek, we use the substrate as the basis of Datnicek (\S\ref{sec:case}), an end-user data
exploration environment inspired by existing systems~\cite{kandel-2011-wrangler,drossos-2020-wrex}.
We use the second stage to evaluate suitability of the Denicek substrate for the development of
end-user programming systems and report the results of our evaluation (\S\ref{sec:eval}).

@@

live/incremental recomputation
\cite{mcdirmid-2013-usable,horowitz-2023-engraft,petricek-2020-live}.

@@

~
% ==================================================================================================

@@

In this section, we use Webnicek, our simple web-based prototype, to illustrate the programming
experiences enabled by the Denicek substrate (\S\ref{sec:impl}). Webnicek is based on a structure
editor that supports navigation in the document and issuing of edit commands. We show two
formative examples (\S\ref{app:examples}) related to conference organization.

\subsubsection*{\circled{A} Adding a Speaker}
The user starts with an empty document, which is represented as a record. They add a field for
each heading and a list {\small\Verb_<ul>_}, represented by a field named {\small\Verb_speakers_}. They use the command
toolbox to add the first speaker.

\subsubsection*{\circled{B} Creating a User Interface} To simplify adding of
further speakers, the user creates a textbox and a button. They enter new speaker
details, add a new {\small\Verb_<li>_} element to the list and copy the speaker details
from the textbox into the new element using source view.

@@

\subsubsection*{\circled{E} Merging Edits} The two document versions can be automatically merged.
The refactoring is applied to all existing speakers. New speakers added using the ``Add speaker!''
button are also automatically added in the new format.

\keyideabox{\faLightbulbO}{Local-First Collaboration}{Denicek's merging
reapplies edits from another branch on top of the current history (\S\ref{sec:impl-collab}).
Merging is asymmetic, but the order does not matter in the above scenario. The same merging
operation is used when handling user interaction (\S\ref{sec:impl-interaction}).}

@@

\keyideabox{\faLightbulbO}{Schema Change Control}{The substrate understands references in
the document and updates them when applying structural edits (\S\ref{sec:impl-schema}).
Evaluation can replace formulas with values, but also augment them to enable
end-user debugging via provenance analysis (\S\ref{sec:impl-provenance}).}

@@

\section{The Denicek Substrate}
\label{sec:system}
Denicek represents programs as sequences of edits that construct and transform a computational
document. In this section, we describe the structure of documents and edits, as well as the
operations that form the backbone of the system and are used to implement the
end-user programming experiences discussed in \S\ref{sec:impl}.

\subsection{Selectors, Documents and Edits}
A computational document is a tree, consisting of four kinds of nodes (Fig.~\ref{fig:doc}).
References to document locations can be relative or absolute. Both kinds can appear in a
document (reference nodes) and absolute references are also used in edits (to denote a target~node).

A reference is represented as a sequence of selectors (Fig.~\ref{fig:doc}).
The document model assumes that lists are homogeneous and records heterogeneous, and so the
\ident{Any} selector makes it possible to refer to all children of a list, but there is no
way to refer to all children of a record. Note that Denicek does not use implicit numerical
indices for lists. The index of a new item has to be supplied explicitly. The design choice
simplifies merging and is discussed in \S\ref{sec:discuss}.

\subsubsection*{Document Edits}
The supported document edits and their behavior are listed in Fig.~\ref{fig:edits}. All edits
require \emph{target} to which they are applied. Target is an absolute reference not containing the
\ident{Parent} selector. It can contain the \ident{Any} selector, in which case the edit is applied
to multiple nodes simultaneously. Most edits can only be applied to target node(s) of a particular
kind.

The edits are designed to allow any transformation of a document
through a series of steps whose effect can be tracked by the substrate. As illustrated earlier,
Denicek updates references when document structure changes. Fig.~\ref{fig:edits} distinguishes
between edits that keep existing references in a document unchanged (above) and edits that affect
references (below).

Renaming a field or wrapping a node require corresponding change to references in the document
(\S\ref{sec:system-ops}). When deleting a field to which there is a reference in the document, Denicek
rejects the edit. A copy edit can also be rejected, because it is ambiguous whether references
referring to the original location should continue referring to the source or should be changed to
point to the target of the copying after the edit (a reference cannot refer to two structurally
independent locations).

@@

\begin{figure}
\newcommand{\ektablecol}[5]{
\raisebox{-0.2em}{#1} & \sffamily\small{\bfseries #2}\,\;\textit{\footnotesize target, #3} & \sffamily\footnotesize #4 \\[-0.2em]
&\multicolumn{2}{l|}{\sffamily\footnotesize #5}\\[0.3em]
}
\begin{tabular}{|cp{17em}r|}
\hline
\rowcolor{ekgray}
&&\\[-1em]
\rowcolor{ekgray}
 & \sffamily\small{\bfseries Edit}\;\,\textit{\footnotesize arguments} & \sffamily\footnotesize\bfseries Target \\[0.2em]
\hline
&&\\[-1em]
\ektablecol{\faPlus}{Add}{field, after, node}{Record}
  {Add \textit{node} as a \textit{field} to the specified record \textit{after} a given field.}
\ektablecol{\faAt}{Append}{index, after, node}{List}
  {Append \textit{node} to the end of the specified list \textit{after} a given field.}
\ektablecol{\faSort}{Reorder}{permutation}{List}
  {Reorder items of a specified list according to a \textit{permutation}.}
\ektablecol{\faMinusCircle}{DeleteItem}{index}{List}
  {Delete the item at a given \textit{index} of a specified list.}
\ektablecol{\faCode}{UpdateTag}{tag}{Tagged}
  {Change the tag of a specified list or record from to a new \textit{tag}.}
\ektablecol{\faICursor}{PrimitiveEdit}{transform}{Primitive}
  {Apply primitive \textit{transform} to the specified primitive.}
\hline
&&\\[-1em]
\ektablecol{\faFont}{RenameField}{old field, new field}{Record}
  {Rename the field of a specified record from \textit{old} to \textit{new}.}
\ektablecol{\faTimesCircle}{DeleteField}{field}{Record}
  {Delete the field \textit{field} of a specified record.}
\ektablecol{\faFileO}{WrapRecord}{tag, field}{Any}
  {Wrap the specified node as a \textit{field} of a new record with \textit{tag}.}
\ektablecol{\faListUl}{WrapList}{tag, index}{Any}
  {Wrap the specified node as a sole element of a new list with \textit{tag}.}
\ektablecol{\faCopy}{Copy}{selectors}{Any}
  {Copy nodes(s) from \textit{selectors}, replacing the specified target(s). }
\hline
\end{tabular}
\vspace{-0.5em}
\caption{Summary of document edit types in Denicek}
\label{fig:edits}
\vspace{-1em}
\end{figure}

@@

\subsubsection*{Automatic Reference Update}
There are two situations in which automatic update of references is undesirable. If an edit is
applied to a singular element of a list (reference contains the \ident{Index} selector),
references that refer into the list should be unchanged. Although such edits violate the invariant
that collections are homogeneous, they typically do so only temporarily, for example during
construction of a new list item (checking such edits is discussed in \S\ref{sec:discuss-future}).

Documents can also contain values that can be of multiple different kinds (i.e., a union type).
In such case, references should not be updated when the kind of the value changes. For example,
a formula may be either unevaluated or evaluated. As discussed in \S\ref{sec:impl-eval}, evaluation
involves wrapping the formula node, but this should not affect references to the formula.
To support those cases, edits that normally affect selectors (Fig.~\ref{fig:edits}, below)
have a \emph{reference behavior} field that can be set to disable automatic reference
updating. This annotation is required when the target contains the \emph{Index} selector.

\subsubsection*{No Conditional Edits.}
An important aspect of the design is that the effect an edit has on references inside
the document does not depend on the current value of the document. This makes it possible to
define merging solely in terms of edits, without reference to current document state.
(The effect \ident{Copy} has on the document structure depends on the existing document
structure, but not on its value).

This design choice makes it impossible to encode computational logic directly in the edits
(e.g., through conditional edits). As we will see in \S\ref{sec:impl-interaction}, such logic
can be provided as an additional mechanism on top of the underlying Denicek substrate.

\subsection{Primitive Operations}
\label{sec:system-ops}
Denicek provides three primitive operations. A sequence of edits can be applied to a document, two
sequences of edits can be merged and also checked for conflicts. Denicek identifies edit histories by
a (git-like) hash, computed from the hash of the parent and the current edit. The hash is used to
identify common shared part of the history during conflict checking and merging.

\subsubsection*{Applying Edits}
When applying an edit, Denicek locates the target node and transform it according to the
edit. If the edit may affect references in the document (Fig~\ref{fig:edits}, below), Denicek
updates matching references in the document according to the rules shown in Fig.~\ref{fig:updates}
provided that the reference behavior is not set to disable the updating (which is required when
the target reference contains the \ident{Index} selector). Although \ident{WrapList} has to specify
the index to be used for the newly created list item, the references in the document are updated to
use the \ident{All} selector. We assume wrapping introduces a homogeneous list and references
pointing to individual items should thus be generalized to point to all list items.

Also note that affected references in the document may be more specific than the edit
target. For example, if we rename {\small\Verb|old|} to {\small\Verb|new|} at {\small\Verb|/foo/*|},
a reference {\small\Verb|/foo/3/old|} will become {\small\Verb|/foo/3/new|}. If the document
contains a reference that would be invalidated by the \ident{Copy} or \ident{DeleteField} edit,
the edit is rejected.

todo

@@

\begin{figure}
\newcommand{\tttablecol}[5]{
\small{\bfseries #1}\;\,\footnotesize\textit{#2}\,\; --\,\; #5\\[-0.1em]
\quad \footnotesize #3 \;\;$\Rightarrow$\;\; #4 \\[0.3em]
}
\begin{tabular}{|l|}
\hline
\\[-1em]
\tttablecol{RenameField}{target, old field, new field}{\Verb|/target/old\_field/nested|}{\Verb|/target/new\_field/nested|}
  {Replace Field for matching references.}
\tttablecol{WrapRecord}{target, tag, field}{\Verb|/target/nested|}{\Verb|/target/field/nested|}
  {Insert extra Field selector after matching prefix.}
\tttablecol{WrapList}{target, index, tag}{\Verb|/target/nested|}{\Verb|/target/*/nested|}
  {Insert extra All selector after matching prefix.}
\hline
\end{tabular}
\vspace{-0.5em}
\caption{How document edits transform references}
\label{fig:updates}
\vspace{-1em}
\end{figure}

@@

\begin{figure}
\newcommand{\extablecol}[5]{
\small{\bfseries #1} #2\,\; --\,\; #3\\[-0.1em]
\quad \footnotesize #4 \\[-0.2em]
\quad \footnotesize \emph{#5} \\[0.4em]
}

\begin{tabular}{|l|}
\hline
\\[-1em]
\extablecol{Counter}{\cite{kiss-2014-7guis}}
  {Counter with increment and decrement buttons.}
  {The current count is represented by a formula that is modified by the buttons.\\[-0.2em]\quad\footnotesize
  The user can inspect the evaluation trace to see how the count was modified.}
  {Programming by Demonstration, Incremental Recomputation, End-User Debugging}
\extablecol{Todo}{\cite{osmani-2024-todomvc}}
  {App with controls to add an item and remove all completed.}
  {Adding an item must correctly merge with independently added functionality to \\[-0.2em]\quad\footnotesize
   compute which items are completed and remove them based on a formula result.}
  {Programming by Demonstration, Local-First Collaboration, Incremental Recomputation}
\extablecol{Conference list}{\cite{edwards-2025-schema}}
  {Manage a list of invited conference speakers.}
  {Adding speakers to a list through an in-document user interface merges with \\[-0.2em]\quad\footnotesize
   refactoring that turns the list into a table and separates name from an email.}
  {Local-First Collaboration, Programming by Demonstration}
\extablecol{Conference budget}{\cite{edwards-2025-schema}}
  {Calculate budget based on a speaker list.}
  {References are updated  when the list is refactored. Only affected formulas\\[-0.2em]\quad\footnotesize
   are recomputed and the user can view elements on which the result depends.}
  {Local-First Collaboration, Incremental Recomputation, End-User Debugging}
\extablecol{Hello world}{\cite{miller-2001-simult}}
  {Normalize the capitalization of two word messages.}
  {An operation to normalize the text in a list item can be recorded and applied to\\[-0.2em]\quad\footnotesize
   all list items or, alternatively, applied directly to all list items.}
  {Programming by Demonstration}
\hline
\end{tabular}

\caption{Structure of selectors and document nodes}
\label{fig:examples}
\vspace{-1em}
\end{figure}

\appendix
\section{Formative Examples}
\label{app:examples}

hello world is only loosely

\cite{osmani-2024-todomvc}
